Research Spike: Evaluation of Using a Single Mnemonic both for Cryptocurrency and Identity Wallets

### 2021-12-21 prism-backend/research/single-mnemonic/discussion.tex
Fixed typo.
Added comments on diffs w/ sec. model of CCS'21 paper, and interoperability.

\paragraph{Domain separation when different cryptosystems are
  necessary.} %
As mentioned, Cardano uses EdDSA (i.e., it is based on Ed25519 curve),
while Atala uses ECDSA with Secp256k1. It is \emph{not} a good idea%
\footnote{I have not been able to find a concrete paper that describes
  some concrete related attack or gives some impossibility result for
  proving security under this circumstance but, at the least, it seems to
  be folklore knowledge. See Lindell's answer at
  \url{https://crypto.stackexchange.com/a/54666/52362} for instance.
  \cite{dlp12+,thorm21} seem good references to study this topic further, if
  needed.} to use the same key for different cryptosystems, nor for different
curves, even though ``structurally'' it may be possible (e.g., in EdDSA and
ECDSA, private keys are 32-byte random numbers). However, this is easily
avoidable by ensuring domain separation in the key derivation functions that
are applied on the common seed. Since this would be handled by the

@@

\paragraph{Careful consideration of extra security properties.} %
The main related work analysed so far\footnote{BIP32, BIP44, \cite{def+21}
  and, partially, \cite{kkl20}.} analyses the security of either BIP44
wallets, or PoS wallets. The introduction of identity-related (frequently
associated to actual people) data may require additional properties, only
mentioned in these works (or not considered, as in the BIPs). For instance,
some sort of forward security will be very desirable. While, for the case of
payments or staking, the impact of a compromise (either of a leaf node --
i.e., a key -- or a complete subtree) can be mitigated by prompt detection
and transfer of funds to an uncompromised address, this may be harder for
keys representing in some way real world identities. Most probably, the
latter will frequently be long-lived keys, that also can, in turn, be used
to issue further identities (e.g., as in the case of identity providers).
Thus, the consequences of a compromise may be more cumbersome to address
(re-issuing credentials, distributing revocation lists, etc.) and,
consequently, it seems desirable to look for constructions that give some
sort of guarantee about the security of keys produced at time $t' < t$, if a
compromise happens at time $t$.

@@

>>>>

\paragraph{Domain separation when different cryptosystems are
  necessary.} %
As mentioned, Cardano uses EdDSA (i.e., it is based on Ed25519 curve),
while Atala uses ECDSA with Secp256k1. It is \emph{not} a good idea%
\footnote{I have not been able to find a concrete paper that describes
  some concrete related attack or gives some impossibility result for
  proving security under this circumstance but, at the least, it seems to
  be folklore knowledge. See Lindell's answer at
  \url{https://crypto.stackexchange.com/a/54666/52362} for instance.
  \cite{dlp12+,thorm21} seem good references to study this topic further, if
  needed.} to use the same key for different cryptosystems, nor for different
curves, even though ``structurally'' it may be possible (e.g., in EdDSA and
ECDSA, private keys are 32-byte random numbers). However, this is easily
avoidable by ensuring domain separation in the key derivation functions that
are applied on the common seed. This is an important concern on its own; yet,
it should be addressed natively through the considerations in the next
paragraph.

@@

\paragraph{Concrete differences in prior security analysis} %
The security analysis in \cite{def+21} is on BIP32, not BIP44. While BIP44
is actually a subset of BIP32, it further specifies it in concrete ways
that may alter the result of the analysis. For instance, \cite{def+21}
assumes that non-hardened nodes are leaves of the tree (which, alternatively,
can be seen as roots in new trees). However, in BIP44 the leaves are
non-hardened nodes, and intermediate nodes are mostly hardened. Yet another
possible source of discrepancy is that the analysis assumes that non-hardened
nodes are maintained in a hot/cold wallet setting and, thus, the private keys
cannot be compromised. This assumption may not hold for all use cases, and
that would change completely the security reasoning. If we want to determine
the concrete security level we want/need in some specific use cases that differ
from the analysis of \cite{def+21} as mentioned above, a new model might be
necessary.

\paragraph{Careful consideration of extra security properties.} %
The main related work analysed so far\footnote{BIP32, BIP44, \cite{def+21}
  and, partially, \cite{kkl20}.} study the security of either BIP32
wallets for payments, or PoS wallets. The introduction of identity-related
(frequently associated to actual people) data may require additional properties,
only mentioned in these works (or not considered, as in the BIPs). For instance,
some sort of forward security will be very desirable. While, for the case of
payments or staking, the impact of a compromise (either of a leaf node --
i.e., a key -- or a complete subtree) can be mitigated by prompt detection
and transfer of funds to an uncompromised address, this may be harder for
keys representing in some way real world identities. Most probably, the
latter will frequently be long-lived keys, that also can, in turn, be used
to issue further identities (e.g., as in the case of identity providers).
Thus, the consequences of a compromise may be more cumbersome to address
(re-issuing credentials, distributing revocation lists, etc.) and,
consequently, it seems desirable to look for constructions that give some
sort of guarantee about the security of keys produced at time $t' < t$, if a
compromise happens at time $t$.

@@

\paragraph{On ``generic'' interoperability issues.} %
Interoperability can mean many things. Hence, in this discussion, we restrict
to the structure of the (hierarchically and deterministically) produced tree,
and on the mnemonic length, which are two differing aspects in Atala and
Cardano (lightwallet). On the length of the mnemonic, Atala uses 12-word
mnemonics, while Cardano uses 24 words. According to BIP39, 12 words encode
128 bits of entropy, and 24 words encode 256. Taking into account that ECDSA
(in the case of Atala) provides a security level of roughly half of the size
of the key, and that the deterministic approach to derive keys from the seed
(HMAC-SHA512) does not amplify entropy \cite{rfc5869}, it is advisable to move
to 24-word mnemonics. Even more, considering the analysis of \cite{def+21},
that concludes that the construction of HD wallets based on additively
re-randomized ECDSA results in a security loss of 31 bits.

Related to the structure of the tree, it is worth mentioning that the Cardano
light wallet uses the conventional BIP44 hierarchy, composed of a 6-levels tree
(including the master seed), where the first four levels are derived in
hardened mode, and the last one in non-hardened mode; on the other hand, Atala
uses a 4-levels tree (again, including the master seed), all of them derived
in hardened mode. Essentially, the Atala wallet ``prunes'' the last two levels.
While the format of the produced keys is of course the same (they are just
random bitstrings), this discrepancy may lead to issues implementation-wise,
that should be further evaluated.
